Tropical mathematics for the Lambda-calculus

### 2023-07-27 CSL24/section3B.tex
Altro commit pre-cena
Commit notturno

As well-known, if a SMCC is \emph{Lafont}, then one obtains a CCC from it by defining the exponential objects as $X\to Y:=!X \multimap Y$.
With no surprise $\LREL$ is Lafont (~\cite[Corollary III.6]{Manzo2013}) %and specialising \cite[Theorem III.7]{Manzo2013} 
w.r.t.\ the usual $!$ acting on objects by taking the set of finite multisets.
So the coKleisli $\LREL_!$ is CCC, i.e.\ a model of $\STLC$.

@@

For what concerns the probabilistic calculus, let us consider a probabilistic extension of $\STLC$, call it $\STLC_\oplus$:
we add a ground type $\bool$, terms $\true,\false$ of type $\bool$, terms of shape $M\oplus_p N$ and $pM$, for $p\in[0,1]$, typed via the usual rules and whose operational semantics adds the usual rules:
{\small{\[\begin{array}{cccc}
           \dfrac{\Gamma\vdash M:A \qquad \Gamma\vdash N:A}{\Gamma\vdash M\oplus_p N:A}
           &
           \dfrac{\Gamma\vdash M:A}{\Gamma\vdash pM:A}
           &
           M\oplus_p N \to pM
           &
           M\oplus_p N \to (1-p)N
          \end{array}\]}}
%We add reduction rules:$ M\oplus_p N \to pM \quad$ and $\quad M\oplus_p N \to (1-p)N,$
so that $M\oplus_p N$ plays the role of a probabilistic coin toss of bias $p$.
We use this calculus as a toy example for our purposes.
It can be compiled into $\mathrm{PCF}^{\Lawv}$, an instance of the $\mathrm{PCF}^{\mathcal R}$-languages studied in \cite{Manzo2013}, via the translation $(\_)^\circ$ of terms generated by $(M\oplus_p N)^\circ:=pM^\circ\prog{ or } (1-p)N^\circ$.
It is easy to see that if $\Gamma \vdash_{\STLC_\oplus} M:A$, then $\Gamma \vdash_{\mathrm{PCF}^{\Lawv}} M^\circ:A$.
This is slightly imprecise because in \cite{Manzo2013}, their $\mathrm{PCF}^{\mathcal R}$-languages only have one ground type $\mathrm{int}$, used for modeling $\N$, and has no term $\prog{True}$, $\prog{False}$.
So we are actually compiling $\STLC_\oplus$ into the immediate extension of $\mathrm{PCF}^{\mathcal R}$ which adds two new ground types, $*$ and $\bool$ (and, since we will not consider the integers, we can also drop $\mathrm{int}$), plus the two boolean terms.
This does not invalidate the results (not concerning $\mathrm{int}$) of \cite{Manzo2013}.
Remark also that the operational semantics of $\mathrm{PCF}^{\Lawv}$ \cite[Fig.\ 1]{Manzo2013} simulates the one of $\STLC_\oplus$, potentially using multiple steps, e.g.\ $(M\oplus_p N)^\circ \overset{0}{\rightarrow} pM^\circ \overset{p}{\rightarrow} M^\circ$.
Since $\LREL_!$ is an instance of models for $\mathrm{PCF}^{\mathcal{R}}$, similarly to \cite[Section VI]{Manzo2013}, we use it in order to get a model of $\STLC_\oplus$, by setting $\model{\Gamma \vdash_{\STLC_\oplus} M:A}:=\model{\Gamma \vdash_{\mathrm{PCF}^{\Lawv}} M^\circ:A}\in\HOM{\LREL_!}{\model{\Gamma}}{\model{A}}$.
Here we set $\model{\bool}:=\{0,1\}$ and $\model{\vdash_{\mathrm{PCF}^{\Lawv}} \true:\bool}\in\Lawv^{\{0,1\}}$ gives value $0$ on $1$ and value $\infty$ on $0$, similarly for $\false$.

Since $\LREL_!$ is an instance of models for $\mathrm{PCF}^{\mathcal{R}}$, similarly to \cite[Section VI]{Manzo2013}, we use it in order to get a model of $\STLC_\oplus$, by setting $\model{\Gamma \vdash_{\STLC_\oplus} M:A}:=\model{\Gamma \vdash_{\mathrm{PCF}^{\Lawv}} M^\circ:A}\in\HOM{\LREL_!}{\model{\Gamma}}{\model{A}}$.
Here we set $\model{\bool}:=\{0,1\}$ and $\model{\vdash_{\mathrm{PCF}^{\Lawv}} \true:\bool}\in\Lawv^{\{0,1\}}$ gives value $0$ on $1$ and value $\infty$ on $0$, similarly for $\false$.

>>>>

As well-known, if a SMCC is \emph{Lafont}, then one obtains a CCC from it by defining the exponential objects as $X\to Y:=!X \multimap Y$.
With no surprise $\LREL$ is Lafont (~\cite[Corollary III.6]{Manzo2013}) %and specialising \cite[Theorem III.7]{Manzo2013} 
w.r.t.\ the usual $!$ acting on objects by taking the set of finite multisets.
So the coKleisli $\LREL_!$ is CCC, i.e.\ a model of $\STLC$.
The cartesian product $\&$ of $\LREL_!$ (and of $\LREL$) is the disjoint union $+$ of sets, the \emph{evaluation} morphisms $\RM{ev}$ are matrices in $\Lawv^{!(!X\times Y)+X)\times Y}$, and the coKleisli composition of $s\in\Lawv^{!Y\times Z}$ and $t\in\Lawv^{!X\times Y}$ is the matrix $s\circ_! t\in\Lawv^{!X\times Z}$, $(s\circ_! t)_{\mu,c}:=
%\begin{align}
\inf_{n\in\N, b_1\dots,b_n\in Y, \mu = \mu_1+\cdots +\mu_n}
 \left\{s_{[b_1,\dots,b_n],c} + \sum_{i=1}^n t_{\mu_i,b_i}\right\}$.

@@

Remember the probabilistic calculus $\STLC_\oplus$.
It can be compiled into $\mathrm{PCF}^{\Lawv}$, an instance of the $\mathrm{PCF}^{\mathcal R}$-languages studied in \cite{Manzo2013}, via the translation $(\_)^\circ$ of terms generated by $(M\oplus_p N)^\circ:=pM^\circ\prog{ or } (1-p)N^\circ$.
It is easy to see that if $\Gamma \vdash_{\STLC_\oplus} M:A$, then $\Gamma \vdash_{\mathrm{PCF}^{\Lawv}} M^\circ:A$.
This is slightly imprecise because in \cite{Manzo2013}, their $\mathrm{PCF}^{\mathcal R}$-languages only have one ground type $\mathrm{int}$, used for modeling $\N$, and has no term $\prog{True}$, $\prog{False}$.
So we are actually compiling $\STLC_\oplus$ into the immediate extension of $\mathrm{PCF}^{\mathcal R}$ which adds two new ground types, $*$ and $\bool$ (and, since we will not consider the integers, we can also drop $\mathrm{int}$), plus the two boolean terms.
This does not invalidate the results (not concerning $\mathrm{int}$) of \cite{Manzo2013}.
Remark also that the operational semantics of $\mathrm{PCF}^{\Lawv}$ \cite[Fig.\ 1]{Manzo2013} simulates the one of $\STLC_\oplus$, potentially using multiple steps, e.g.\ $(M\oplus_p N)^\circ \overset{0}{\rightarrow} pM^\circ \overset{p}{\rightarrow} M^\circ$.
Since $\LREL_!$ is an instance of models for $\mathrm{PCF}^{\mathcal{R}}$, similarly to \cite[Section VI]{Manzo2013}, we use it in order to get a model of $\STLC_\oplus$, by setting $\model{\Gamma \vdash_{\STLC_\oplus} M:A}:=\model{\Gamma \vdash_{\mathrm{PCF}^{\Lawv}} M^\circ:A}\in\HOM{\LREL_!}{\model{\Gamma}}{\model{A}}$.
Here we set $\model{\bool}:=\{0,1\}$ and $\model{\vdash_{\mathrm{PCF}^{\Lawv}} \true:\bool}\in\Lawv^{\{0,1\}}$ gives value $0$ on $1$ and value $\infty$ on $0$, similarly for $\false$.

### 2023-07-27 CSL24/section4.tex
Commit notturno

\begin{theorem}\label{thm:cont}
 All tLs $f:\Lawv^X\to\Lawv$ are continuous on $(0,\infty)^X$, w.r.t.\ to the norm $\norm{\cdot}_\infty$.
\end{theorem}
\begin{proof}
%It follows after adapting \cite[Proposition 4.4]{Cobzas2017} in order to prove that if a real-valued function on a locally convex topological $\BB R$-vector space is, locally around $x$, concave and bounded by a finite constant, then it is continuous at $x$.
\cite[Proposition 4.4.(3)]{Cobzas2017} shows that if a real-valued convex function with domain a convex open subset of a locally convex topological $\BB R$-vector space is, locally around any point, bounded from above by a finite non-zero constant, then it is continuous on all its domain.
Now, our $f$ is concave, so $-f$ is convex.
Also, since $f\geq 0$ on all $\Lawv^X$, we have e.g.\ $-f\leq 1$ on $\Lawv^X$.
%Remark that we cannot immediately apply \cite[Proposition 4.4.(3)]{Cobzas2017}, since $\Lawv^X$ is not even a $\BB R$-vector space.
Now $(0,\infty)^X\subseteq \BB R^X$ is open and convex, and {\color{red}$\BB R^X$ is a locally convex topological $\BB R$-vector space}, so \cite[Proposition 4.4.(3)]{Cobzas2017} entails the continuity of $-f\big|_{(0,\infty)^X}$, hence that of $f\big|_{(0,\infty)^X}$.
\end{proof}

>>>>

\begin{theorem}\label{thm:cont}
 All tLs $f:\Lawv^X\to\Lawv$ are continuous on $(0,\infty)^X$, w.r.t.\ to the norm $\norm{\cdot}_\infty$.
\end{theorem}
\begin{proof}
%It follows after adapting \cite[Proposition 4.4]{Cobzas2017} in order to prove that if a real-valued function on a locally convex topological $\BB R$-vector space is, locally around $x$, concave and bounded by a finite constant, then it is continuous at $x$.
\cite[Proposition 4.4.(3)]{Cobzas2017} shows that if a real-valued convex function with domain a convex open subset of a locally convex topological $\BB R$-vector space (LCTVS) is, locally around any point, bounded from above by a finite non-zero constant, then it is continuous on all its domain.
Now, our $f$ is concave, so $-f$ is convex.
Also, since $f\geq 0$ on all $\Lawv^X$, we have e.g.\ $-f\leq 1$ on $\Lawv^X$.
%Remark that we cannot immediately apply \cite[Proposition 4.4.(3)]{Cobzas2017}, since $\Lawv^X$ is not even a $\BB R$-vector space.
Now $(0,\infty)^X\subseteq \BB R^X$ is open and convex, and {\color{red}$\BB R^X$ is a LCTVS}, so \cite[Proposition 4.4.(3)]{Cobzas2017} entails the continuity of $-f\big|_{(0,\infty)^X}$, hence that of $f\big|_{(0,\infty)^X}$.
\end{proof}

### 2023-07-27 CSL24/section5.tex
Commit notturno

Consider %the following closed term $M$ of type $\bool$:
$
 \vdash_{\STLC_\oplus} M:=(\true \oplus_p\false)\oplus_p((\true\oplus_p\false)\oplus_p(\false\oplus_p\true)):\bool.
 $
Let us give addresses $\omega\in\set{00,01,100,101,110,111}$ to the occurrences of $\true,\false$ in $M$, by following the tree structure of $M$, reading $0$ as ``left'' and $1$ as ``right''.
The same addresses also represent all the different possible reduction paths from $M$ to a normal form.
%For instance, $100$ represents the reduction which keeps the right part of the outermost $\oplus_p$ and erases the left part, then continues by choosing the left part twice, reaching at the end the occurrence $\true_{100}$ in $M$, i.e.\ the second occurrence of $\true$ in $M$ starting from the left.
Calling $q:=1-p$, there are the following six normal terms reachable from $M$:
$P_{00}(p,q)\true$, 
$P_{100}(p,q)\true$, 
$P_{111}(p,q)\true$, 
$P_{01}(p,q)\false$, 
$P_{110}(p,q)\false$,
$P_{101}(p,q)\false$,
where the $P$'s are the following monomials in $p,q$:
$P_{00}(p,q):=p^2$,
$P_{100}(p,q):=qp^2$,
$P_{111}(p,q):=q^3$,
$P_{01}(p,q):=pq$,
$P_{110}(p,q)=P_{101}(p,q):=q^2p$.
They correspond to the respective reduction path from $M$ to the normal term of the same address.
$P_{\omega}$ is then the probability (as a function of $p,q$) of obtaining the respective occurrence $\true_{\omega}$ or $\false_\omega$ after all the coin tossings in $M$ are performed.
Thinking of $p,q$ as parameters, $P_{\omega}$ can thus be read as the \emph{likelihood function} of the event $\omega$.
The polynomial $Q_{1}(p,q):=P_{00}(p,q)+P_{100}(p,q)+P_{111}(p,q)$ gives instead the whole probability of obtaining $\true$ after all the tossings, and $Q_{0}(p,q):=P_{01}(p,q)+P_{110}(p,q)+P_{101}(p,q)$ the one for $\false$.

>>>>

For what concerns the probabilistic calculus, let us consider a probabilistic extension of $\STLC$, call it $\STLC_\oplus$:
we add a ground type $\bool$, terms $\true,\false$ of type $\bool$, terms of shape $M\oplus_p N$ and $pM$, for $p\in[0,1]$, typed via the usual rules and whose operational semantics adds the usual rules:
{\small{\[\begin{array}{cccc}
           \dfrac{\Gamma\vdash M:A \qquad \Gamma\vdash N:A}{\Gamma\vdash M\oplus_p N:A}
           &
           \dfrac{\Gamma\vdash M:A}{\Gamma\vdash pM:A}
           &
           M\oplus_p N \to pM
           &
           M\oplus_p N \to (1-p)N
          \end{array}\]}}
%We add reduction rules:$ M\oplus_p N \to pM \quad$ and $\quad M\oplus_p N \to (1-p)N,$
so that $M\oplus_p N$ plays the role of a probabilistic coin toss of bias $p$.
We use this calculus as a toy example for our purposes.
Consider %the following closed term $M$ of type $\bool$:
$
 \vdash_{\STLC_\oplus} M:=(\true \oplus_p\false)\oplus_p((\true\oplus_p\false)\oplus_p(\false\oplus_p\true)):\bool.
 $
Let us give addresses $\omega\in\set{00,01,100,101,110,111}$ to the occurrences of $\true,\false$ in $M$, by following the tree structure of $M$, reading $0$ as ``left'' and $1$ as ``right''.
The same addresses also represent all the different possible reduction paths from $M$ to a normal form.
%For instance, $100$ represents the reduction which keeps the right part of the outermost $\oplus_p$ and erases the left part, then continues by choosing the left part twice, reaching at the end the occurrence $\true_{100}$ in $M$, i.e.\ the second occurrence of $\true$ in $M$ starting from the left.
Calling $q:=1-p$, there are the following six normal terms reachable from $M$:
$P_{00}(p,q)\true$, 
$P_{100}(p,q)\true$, 
$P_{111}(p,q)\true$, 
$P_{01}(p,q)\false$, 
$P_{110}(p,q)\false$,
$P_{101}(p,q)\false$,
where the $P$'s are the following monomials in $p,q$:
$P_{00}(p,q):=p^2$,
$P_{100}(p,q):=qp^2$,
$P_{111}(p,q):=q^3$,
$P_{01}(p,q):=pq$,
$P_{110}(p,q)=P_{101}(p,q):=q^2p$.
They correspond to the respective reduction path from $M$ to the normal term of the same address.
$P_{\omega}$ is then the probability (as a function of $p,q$) of obtaining the respective occurrence $\true_{\omega}$ or $\false_\omega$ after all the coin tossings in $M$ are performed.
Thinking of $p,q$ as parameters, $P_{\omega}$ can thus be read as the \emph{likelihood function} of the event $\omega$.
The polynomial $Q_{1}(p,q):=P_{00}(p,q)+P_{100}(p,q)+P_{111}(p,q)$ gives instead the whole probability of obtaining $\true$ after all the tossings, and $Q_{0}(p,q):=P_{01}(p,q)+P_{110}(p,q)+P_{101}(p,q)$ the one for $\false$.
